---
title: "Maximum likelihood estimation (MLE) of tuskless elephant survival bias"
output: html_document
---
library(BiasedUrn)
### Raw Data

total pre-war popuation size:      2542  
total post-war population size:    242  
female tuskless proportion before: 0.185  
female tuskless proportion after:  0.509

These population size estimates are for the entirety of Gorongosa, but tusk status was only studied
in females since they are the only sex that are polymorphic for the presence of tusks. We thus focus on the number of female elephants before and after the war to quantify the survival bias of tuskless females compared to tusked females. 

While we do not know precisely how many of these elephants before and after the war are females, a 50:50 sex ratio is likely a sensible lower bound because (1) humans may preferentially target males because of their
larger tusks and (2) if tusklessness is a male-lethal trait, this will skew the population slightly towards females.

```{r, echo=FALSE }
sexRatio <- 0.5

Npre <- 2542*sexRatio # total population size of females before war
Npost <- 242*sexRatio # total population size of females after war
PropTuskless_pre <- 0.185 # fraction of tuskless females before war
PropTuskless_post <- 0.509 # fraction of tuskless females after war

Npre_tuskless <- Npre*PropTuskless_pre # number of tuskless females before the war
Npre_tusked <- Npre*(1-PropTuskless_pre) # number of tusked females before the war
Npost_tuskless <- Npost*PropTuskless_post # number of tuskless females after the war
Npost_tusked <- Npost*(1-PropTuskless_post) # number of tusked females after the war
```


### Computing the MLE

To model the survival bias of tuskless female elephants, we use the Wallenius distribution that is featured in the BiasedUrn package ("dWNCHypergeo" function). This distribution models the subsampling of a population containing two distinct types.

```{r, echo=FALSE }
library(scales)
library(BiasedUrn)
```

This is the null distribution of how many tuskless females we would expect after the war if there were no difference in survival between tuskless and tusked females.

```{r, echo=FALSE }
x <- rWNCHypergeo(nran=10000000,
               m1=Npre_tuskless,
               m2=Npre_tusked,
               n=(Npost_tuskless+Npost_tusked),
               odds=1) # diff odds ratios input here
hist(x/(Npost_tuskless+Npost_tusked),
     freq=F,
     breaks=20,
     main="",
     xlab="Proportion of Tuskless Females",
     xlim=c(0,0.6),
     col=alpha("#2c7fb8", 0.8))
abline(v=quantile(x/(Npost_tuskless+Npost_tusked), probs=0.99), lty=3)
abline(v=PropTuskless_post)
abline(v=0.3614458)

```

From the documentation:
```
dWNCHypergeo(x, m1, m2, n, odds, precision=1E-7)
x: number red balls sampled
m1: initial number of red balls
m2: initial number of white balls
n: total number of balls sampled
odds: probability ratio of red over white balls
```



Here we treat red balls as tuskless females and white balls as tusked females. Using the data above, we have estimates for all parameters except 'odds'. We will use this dWNCHypergeo functions to find the value of 'odds' that maximizes the likelihood of our data before and after the war by computing the likelihood for a series of 'odds' values ranging from no bias (odds=1) to extreme bias (odds=10). We assume the true bias lies somewhere within this interval, and how densely we sample points within this range will affect the accuracy of our MLE estimates and confidence intervals (CIs).
```{r, echo=FALSE }
odds_range <- seq(from=1, to=10, by=0.001)
```


We will calculate the likelihood for each one of these 'odds' values and store them in a vector.
```{r, echo=FALSE }
likelihoods <- c()
for(i in 1:length(odds_range)){
  tmp <- dWNCHypergeo(x=Npost_tuskless,
               m1=Npre_tuskless,
               m2=Npre_tusked,
               n=(Npost_tuskless+Npost_tusked),
               odds=odds_range[i]) # diff odds ratios input here
  likelihoods <- c(likelihoods, tmp) # store likelihoods as we go
}
# the maximum likelihood value
max(likelihoods)
# the parameter 'odds' value that corresponds to this maximum likelihood value (MLE of 'odds')
odds_range[ likelihoods == max(likelihoods) ] 
```

A plot of these likelihoods:  
```{r, echo=FALSE }
library(scales)
plot(y=likelihoods, x=odds_range,
     type="l",
     lwd=3,
     axes=F,
     xlab="",
     ylab="",
     xlim=c(2,10),
     main="likelihood")
axis(side=1, at=c(2:10), labels=T)
axis(side=2, at=seq(from=0, to=0.06, by=0.01))
mtext(side=1, text="tuskless survival bias", line=2)
mtext(side=2, text="likelihood", line=2)
abline(v= odds_range[ likelihoods == max(likelihoods) ], lty=2, col=alpha("black", 0.5) )
```


We will repeat the above process except use log likelihoods, which we may use to compute CIs since a drop in 1.92 log likelihoods from the maximum value corresponds to the 95% CI.
```{r, echo=FALSE }
loglikelihoods <- c()
for(i in 1:length(odds_range)){
  tmp <- dWNCHypergeo(x=Npost_tuskless,
                      m1=Npre_tuskless,
                      m2=Npre_tusked,
                      n=(Npost_tuskless+Npost_tusked),
                      odds=odds_range[i])
  tmp <- log(tmp)
  loglikelihoods <- c(loglikelihoods, tmp)
}

# The MLE of 'odds', should be exact same as above
MLE <- odds_range[ loglikelihoods == max(loglikelihoods) ] 
# To compute 95% CIs, get all 'odds' that are within 1.92 log likelihood units from the MLE. Note: the following vector just contains indices
idx_mleCI <- which( loglikelihoods >= (max(loglikelihoods) - 1.92)  )
# Use the minimum and maximum value of this vector of indices to get the endpoints of the CI, or the 95% CI.
CI <- odds_range[ c(min(idx_mleCI), max(idx_mleCI)) ]
```

A plot of these log likelihoods and the 95% CI of our MLE:  
```{r, echo=TRUE }
dev.new(height=5, width=5, noRStudioGD=T)
plot(y=(loglikelihoods), x=odds_range,
     lwd=3,
     type="l",
     axes=F,
     xlab="",
     ylab="",
     xlim=c(3,8),
     ylim=c(-8,-2),
     col=alpha("#2c7fb8", 0.8))
axis(side=1, labels=T, at=c(1:10))
axis(side=2, labels=T)
mtext(side=1, text="tuskless survival bias", line=2)
mtext(side=2, text="log(likelihood)", line=2)
#abline(h=(max(loglikelihoods) - 1.92), lty=2, col=alpha("black", 0.5))
abline(v=MLE, col=alpha("black", 0.8), lty=3, lwd=3)
abline(v=CI, col=alpha("red", 0.5), lty=3, lwd=3)
```


Multipanel figure for the manuscript

```{r, echo=TRUE }
dev.new(height=7, width=5, noRStudioGD=T)
par(mfrow=c(2,1), mar=c(4,4,1,1))
x <- rWNCHypergeo(nran=10000000,
               m1=Npre_tuskless,
               m2=Npre_tusked,
               n=(Npost_tuskless+Npost_tusked),
               odds=1) # diff odds ratios input here
hist(x/(Npost_tuskless+Npost_tusked),
     freq=F,
     breaks=20,
     main="",
     xlab="",
     ylab="",
     xlim=c(0,0.6),
     col=alpha("#2c7fb8", 0.8))
mtext(side=2, text="Density", line=2.5)
mtext(side=1, text="proportion of tuskless females", line=2)
abline(v=quantile(x/(Npost_tuskless+Npost_tusked), probs=0.99), lty=3)
abline(v=PropTuskless_post)
abline(v=0.3614458)

plot(y=(loglikelihoods), x=odds_range,
     lwd=3,
     type="l",
     axes=F,
     xlab="",
     ylab="",
     xlim=c(3,8),
     ylim=c(-8,-2),
     col=alpha("#2c7fb8", 0.8))
axis(side=1, labels=T, at=c(1:10))
axis(side=2, labels=T)
mtext(side=1, text="tuskless survival odds", line=2)
mtext(side=2, text="log(likelihood)", line=2.5)
#abline(h=(max(loglikelihoods) - 1.92), lty=2, col=alpha("black", 0.5))
abline(v=MLE, col=alpha("black", 0.8), lty=3, lwd=3)
abline(v=CI, col=alpha("red", 0.5), lty=3, lwd=3)
```
